---
title: "Bio-logger Data from Ribbon and Spotted Seals: A Deep Dive"
author: "Josh M. London"
format: 
  revealjs:
    smaller: true
---

## Introduction

:::: {.columns}

::: {.column width="50%"}
[insert family photo]
:::

::: {.column width="50%"}
**Josh**

- Originally from Tulsa, Oklahoma
- Seattle since 1993
- PhD from Univ. of Washington (seals, salmon, killer whales)

**Shanna**

- Originally from Mexico City
- Seattle since 1994

:::

::::

## What are Bio-loggers?

:::: {.columns}

::: {.column width="40%"}
[insert photo bio-logger on seal]
:::

::: {.column width="60%"}
- Smart-phone/Smart-watch for Seals 
  - collect data on location and behavior
- Must Operate in Harsh Conditions 
  - deep dives, cold temperatures
- Battery Life is Limited 
  - no option for seals to plug-in & re-charge
- Data Transfer via Satellite 
  - no WiFi on ice floes
:::

::::

## Attaching Bio-loggers to Seals

#### Methods
- Adhere to seal hair with super glue or epoxy
- Attach to rear-flipper webbing for long-term

#### Ethics and Humaneness

- Marine Mammal Protection Act - Research Permit
- US Animal Welfare Act - Institutional Review (IACUC)

**Key Principle: Maximize Information Gained with Least Impact**

## Bio-logger Information: Ribbon and Spotted Seals

1. Movement and Space Use
2. Dive Behavior
3. Haul-out Behavior
4. Oceanography

## Movement and Space Use

## Dive Behavior of Ribbon Seals

```{r ribbon-tad-plot, warning=FALSE, message = FALSE}
library(tidyverse)
library(dbplyr)
library(DBI)
con <- dbConnect(odbc::odbc(), dsn = "PostgreSQL pep", 
                 uid = keyringr::get_kc_account("pgpep_londonj"), 
                 pwd = keyringr::decrypt_kc_pw("pgpep_londonj"))

qry <- "SELECT a.deployid deployid,a.tad_start_dt,EXTRACT(YEAR FROM a.tad_start_dt) deploy_year, a.bin,a.pct_tad, 
          b.deploy_dt deploy_dt, b.end_dt end_dt, b.species species
        FROM telem.tbl_wc_histos_timeatdepth_qa a
        LEFT JOIN telem.tbl_tag_deployments b on a.deployid = b.deployid
        WHERE species IN ('Hf','Pl') AND 
          (a.tad_start_dt > b.deploy_dt and a.tad_start_dt < b.end_dt)"

lku_binlimits_2007 <- tibble::tribble(
  ~bin, ~bin_upper_limit,
  "bin1",     4,
  "bin2",     10,
  "bin3",     30,
  "bin4",     50,
  "bin5",     70,
  "bin6",     90,
  "bin7",     110,
  "bin8",     130,
  "bin9",     150,
  "bin10",    200,
  "bin11",    250,
  "bin12",    300,
  "bin13",    400,
  "bin14",    600
)

lku_binlimits_2009 <- tibble::tribble(
  ~bin, ~bin_upper_limit,
  "bin1",     4,
  "bin2",     10,
  "bin3",     30,
  "bin4",     50,
  "bin5",     70,
  "bin6",     90,
  "bin7",     100,
  "bin8",     150,
  "bin9",     200,
  "bin10",    250,
  "bin11",    300,
  "bin12",    400,
  "bin13",    500,
  "bin14",    600
)

tad_data_hf_a <- tbl(con, from = sql(qry)) %>% 
  dplyr::filter(species == 'Hf',
                deploy_year < 2009) %>% 
  collect() %>% 
  left_join(lku_binlimits_2007, by = "bin")

tad_data_hf_b <- tbl(con, from = sql(qry)) %>% 
  dplyr::filter(species == 'Hf',
                deploy_year > 2008) %>% 
  collect() %>% 
  left_join(lku_binlimits_2009, by = "bin")

tad_data_hf <- rbind(tad_data_hf_a, tad_data_hf_b) %>% 
  dplyr::filter(!is.na(pct_tad)) %>% 
  dplyr::filter(lubridate::hour(tad_start_dt) %in% c(3,9,15,21)) %>% 
  dplyr::mutate(doy_dt = lubridate::yday(tad_start_dt) + 
                  lubridate::hour(tad_start_dt)/24) %>% 
  dplyr::group_by(doy_dt, bin_upper_limit) %>% 
  dplyr::summarise(median_pct_tad = median(pct_tad, na.rm=TRUE)) %>% 
  dplyr::filter(median_pct_tad > 0) %>% 
  dplyr::arrange(doy_dt, bin_upper_limit) %>% 
  dplyr::mutate(prev_depth = dplyr::lag(bin_upper_limit)) %>% 
  dplyr::mutate(prev_depth = case_when(is.na(prev_depth) ~ 0,
                                       TRUE ~ prev_depth)) %>% 
  dplyr::mutate(bin_upper_limit = bin_upper_limit * -1,
                prev_depth = prev_depth * -1)

library(MetBrewer)

ggplot(tad_data_hf) +
geom_rect(aes(xmin = doy_dt-0.125, xmax = doy_dt+0.125,
              ymin = bin_upper_limit, ymax = prev_depth,
              fill = median_pct_tad),
          size=0) +
  scale_fill_gradientn(colors=met.brewer("Tam", direction = -1),
                       guide = guide_colorbar(
                         title = 'Percent Time',
                         title.position = 'bottom',
                         title.hjust = 0.5,
                         barwidth = unit(75, units = "mm"),
                         barheight = unit(2, units = "mm"))) +
  scale_x_continuous(breaks = c(1, 32, 60, 91, 121, 152, 182, 214, 245, 276, 307,338),
                     labels = c("01 Jan","01 Feb", "01 Mar", "01 Apr", "01 May", "01 Jun", "01 Jul",
                                "01 Aug","01 Sep","01 Oct","01 Nov", "01 Dec"),
                     expand = expansion(mult = c(0.02, 0))) +
  xlab("") + ylab("depth (meters)") +
  labs(
    title = "Seasonal Changes in Percent Time at Depth for Ribbon Seals",
    subtitle = "median percent time at depth across multiple bio-logger deployments from 2005-2016",
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
## Dive Behavior of Spotted Seals

```{r spotted-tad-plot, warning=FALSE, message = FALSE}
tad_data_pl_a <- tbl(con, from = sql(qry)) %>% 
  dplyr::filter(species == 'Pl',
                deploy_year < 2009) %>% 
  collect() %>% 
  left_join(lku_binlimits_2007, by = "bin")

tad_data_pl_b <- tbl(con, from = sql(qry)) %>% 
  dplyr::filter(species == 'Pl',
                deploy_year > 2008) %>% 
  collect() %>% 
  left_join(lku_binlimits_2009, by = "bin")

tad_data_pl <- rbind(tad_data_pl_a, tad_data_pl_b) %>% 
  dplyr::filter(!is.na(pct_tad)) %>% 
  dplyr::filter(lubridate::hour(tad_start_dt) %in% c(3,9,15,21)) %>% 
  dplyr::mutate(doy_dt = lubridate::yday(tad_start_dt) + 
                  lubridate::hour(tad_start_dt)/24) %>% 
  dplyr::group_by(doy_dt, bin_upper_limit) %>% 
  dplyr::summarise(median_pct_tad = median(pct_tad, na.rm=TRUE)) %>% 
  dplyr::filter(median_pct_tad > 0) %>% 
  dplyr::arrange(doy_dt, bin_upper_limit) %>% 
  dplyr::mutate(prev_depth = dplyr::lag(bin_upper_limit)) %>% 
  dplyr::mutate(prev_depth = case_when(is.na(prev_depth) ~ 0,
                                       TRUE ~ prev_depth)) %>% 
  dplyr::mutate(bin_upper_limit = bin_upper_limit * -1,
                prev_depth = prev_depth * -1)

ggplot(tad_data_pl) +
geom_rect(aes(xmin = doy_dt-0.125, xmax = doy_dt+0.125,
              ymin = bin_upper_limit, ymax = prev_depth,
              fill = median_pct_tad),
          size=0) +
  scale_fill_gradientn(colors=met.brewer("Tam", direction = -1),
                       guide = guide_colorbar(
                         title = 'Percent Time',
                         title.position = 'bottom',
                         title.hjust = 0.5,
                         barwidth = unit(75, units = "mm"),
                         barheight = unit(2, units = "mm"))) +
  scale_x_continuous(breaks = c(1, 32, 60, 91, 121, 152, 182, 214, 245, 276, 307,338),
                     labels = c("01 Jan","01 Feb", "01 Mar", "01 Apr", "01 May", "01 Jun", "01 Jul",
                                "01 Aug","01 Sep","01 Oct","01 Nov", "01 Dec"),
                     expand = expansion(mult = c(0.02, 0))) +
  xlab("") + ylab("depth (meters)") +
  labs(
    title = "Seasonal Changes in Percent Time at Depth for Spotted Seals",
    subtitle = "median percent time at depth across multiple bio-logger deployments from 2005-2018",
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 2022 - Ribbon Seal Mother w/ Pup

```{r ribbon-mom-plot, warning=FALSE, message = FALSE}
library(wcECDF)

library(here)

ecdhisto <- "~/downloads/dys2022/dys2022-ECDHistos.csv" 

dys2022_TAD <- ecdhisto %>% wcECDF::read_ecdf() %>% 
  dplyr::filter(deploy_id %in% c("HF2022_1005_19U3070")) %>%
  dplyr::mutate(facet_title = case_when(
    deploy_id == "HF2022_1005_19U3070" ~ "Ribbon Seal Adult Female - 2022"
  )) %>% 
  dplyr::arrange(deploy_id, start) %>% 
  dplyr::select(c(deploy_id,facet_title,start:percent_dry,shallow_ecdf, deep_ecdf,full_pdf)) %>% 
  unnest(full_pdf, keep_empty = TRUE) %>% 
  group_by(deploy_id, start, end, kind, percent_dry) %>% 
  dplyr::arrange(deploy_id, start, depth_break) %>% 
  # determine the next depth value
  dplyr::mutate(next_depth = dplyr::lead(depth_break)) %>%
  # determine previous depth for plotting
  dplyr::mutate(prev_depth = dplyr::lag(depth_break)) %>%
  dplyr::select(-next_depth) %>%
  # transform depths to negative values
  dplyr::mutate(depth_break = depth_break * -1,
                prev_depth = prev_depth * -1)

ggplot(dys2022_TAD) +
  geom_rect(aes(xmin = start, xmax = end,
                ymin = 0, ymax = percent_dry/10),
            fill = "seashell3") +
  geom_rect(aes(xmin = start, xmax = end,
                ymin = depth_break, ymax = prev_depth,
                fill = 60*(1 - percent_dry*0.01)*prob),
            color = "white",size=0.25) +
  scale_fill_gradientn(colors=met.brewer("Tam", direction = -1),
                         guide = guide_colorbar(
                         title = 'Minutes per Hour',
                         title.position = 'bottom',
                         title.hjust = 0.5,
                         barwidth = unit(75, units = "mm"),
                         barheight = unit(2, units = "mm"))) +
  facet_wrap(~ facet_title, nrow=2) +
  theme_minimal() +
  ylab("depth (meters)") +
  labs(title = "Dive Behavior of a Ribbon Seal Mom with Pup",
          subtitle = "Hourly distribution of minutes spent at depth",
          caption = "data derived from the Emperical Cummulative Distribution (ECD) of time at depth") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

## Haul-out Behavior of Ribbon and Spotted Seals

## Oceanographic Data

